---
title: "Lab: Ridge Regression and the Lasso"
author: "Reto Wüest"
date: "July 8, 2019"
output:
  html_document:
    highlight: textmate
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data

We will use data from the 2009 British Social Attitudes Survey. You can download the data [here](bsas_short.RData). The codebook is available [here](bsas_doc.pdf).

Our goal is to predict the following output variable (the name in brackets corresponds to the variable name in the codebook):

- **imm_brit** [IMMBRIT]: Out of every $100$ people in Britain, how many do you think are immigrants from non-Western countries?

We will use the following input variables:

- **resp_female** [RSex]: Is the respondent female?
- **resp_age** [RAge]: Age of the respondent
- **resp_household_size** [Househld]: How many people live in respondent's household?
- **resp_party_cons** [PartyIDN]: Respondent is most likely to support the Conservative Party
- **resp_party_lab** [PartyIDN]: Respondent is most likely to support the Labour Party
- **resp_party_libdem** [PartyIDN]: Respondent is most likely to support the Liberal Democratic Party
- **resp_party_snp** [PartyIDN]: Respondent is most likely to support the Scottish National Party
- **resp_party_green** [PartyIDN]: Respondent is most likely to support the Green Party
- **resp_party_ukip** [PartyIDN]: Respondent is most likely to support the UK Independence Party
- **resp_party_bnp** [PartyIDN]: Respondent is most likely to support the British National Party
- **resp_party_other** [PartyIDN]: Respondent is most likely to support another party, no party, refused to say, or did not know
- **resp_newspaper** [Readpap]: Respondent normally reads a daily newspaper
- **resp_internet_hrs** [WWWHrsWk]: Hours the respondent spends using the internet (per week)
- **resp_religious** [Religion]: Respondent regards himself/herself belonging to a particular religion
- **resp_time_current_employment** [EmploydT]: Months respondent has spent with current employer
- **resp_urban_area** [PopBand]: Population density
- **resp_health** [SRHealth]: Respondent's health
- **resp_household_income** [HHincome]: Total income of respondent's household

Let's have a look at the distribution of the output variable. The red dashed line shows the true percentage of non-Western immigrants (see [www.migrationwatchuk.org](https://www.migrationwatchuk.org)).

```{r, echo=FALSE, message=FALSE}
library(foreign)
library(dplyr)
library(ggplot2)

load("bsas_short.RData")

response_count <- bsas_data %>%
  dplyr::group_by(imm_brit) %>%
  dplyr::summarise(no_imm_count = n()) %>%
  data.frame()

nw_immi <- 100 * (6.973 / 65.14)  # True percentage of non-western immigrants in the UK

bar_plot <- ggplot(data = response_count,
                   aes(x = imm_brit, y = no_imm_count)) +
  geom_vline(xintercept = nw_immi, colour = "red", linetype = "dashed") +
  geom_bar(stat = "identity", position = position_dodge()) +
  geom_text(aes(label = no_imm_count),
            position = position_dodge(width = 0.9), vjust = -0.5, size = 2) +
  scale_x_continuous(name = "Estimated percentage of non-Western immigrants") +
  scale_y_continuous(name = "Number of respondents") +
  theme_bw() +
  theme(axis.text.x = element_text(size = 10, margin = margin(t = 5, b = 5)),
        axis.text.y = element_text(size = 10, margin = margin(l = 5, r = 5)),
        axis.title.x = element_text(size = 12),
        axis.title.y = element_text(size = 12))

bar_plot
```

The plot shows that 756 of 1,044 respondents (or 72.4\%) overestimate the percentage of non-Western immigrants in the UK.

## Data Preparation

The original data set contains 1,044 respondents. 

```{r, echo=TRUE, message=FALSE}
library(dplyr)

# Set working directory
# setwd(...)

# Load data set
load("bsas_short.RData")

# Declare factor variables
bsas_data <- bsas_data %>%
  dplyr::mutate(resp_urban_area = factor(resp_urban_area,
                                         levels = 1:4,
                                         labels = c("rural", "rather rural",
                                                    "rather urban", "urban")),
                resp_health = factor(resp_health,
                                     levels = 0:3,
                                     labels = c("bad", "fair", "fairly good", "good")))
```

We will use the `glmnet()` function in the `glmnet` package to perform ridge regression and the lasso. Before doing so, we use the `model.matrix()` function to create a matrix of input variables (also called "design matrix"). This function automatically transforms any qualitative variables into dummy variables. This is important because `glmnet()` can only take quantitative inputs.

We remove the intercept from the matrix produced by `model.matrix()` because `glmnet()` will automatically include an intercept. We also exclude the input **resp_party_cons**, which will serve as the baseline in our model.

```{r, echo=TRUE}
# Matrix of input variables (remove the intercept and resp_party_cons)
x <- model.matrix(imm_brit ~ . -1 -resp_party_cons, bsas_data)

# Output variable
y <- bsas_data$imm_brit
```

## Exercise 1: Ridge Regression

The goal of the first exercise is to fit a ridge regression to our data.

### Problem 1.1

a. Split the data into a training set and a test set.
b. Define a vector of values for the tuning parameter $\lambda$. Use the training set to fit a ridge regression for each $\lambda$ value. You can do this by using the `glmnet()` function in the `glmnet` package. The `glmnet()` function has an `alpha` argument that determines what type of model is fit. If `alpha = 0`, then a ridge regression model is fit, and if `alpha = 1`, then a lasso model is fit.
c. Use the `plot()` function to create a graph that shows the shrinkage of the coefficient estimates as a function of $\lambda$.

### Problem 1.2

a. Perform cross-validation on the trainig set to choose the optimal tuning parameter $\lambda$. You can do this by using the cross-validation function `cv.glmnet()` in the `glmnet` package. Use the `plot()` function to create a graph that shows the CV estimate of the expected test MSE associated with each value of $\lambda$.
b. Find the optimal value of $\lambda$. Use the test set to compute the test error associated with the optimal model.
c. Re-fit the ridge regression model on the full data set, using the optimal value of $\lambda$.

## Exercise 2: The Lasso

The goal of the second exercise is to fit a lasso model to our data.

### Problem 2.1

a. Use the training set to fit a lasso model for each $\lambda$ value. You can again use the `glmnet()` function, setting the argument `alpha = 1`. Use the `plot()` function to create a graph that shows the shrinkage of the estimated coefficients as a function of $\lambda$.
b. Perform cross-validation on the training set to find the optimal value of $\lambda$. Plot the CV error associated with each value of $\lambda$ and use the test set to compute the test error for the optimal model.
c. Re-fit the lasso model on the full data set, using the optimal value of $\lambda$.
